---
title: "RUMT Dplyr"
author: "Josh Nowak"
date: "September 30, 2015"
output: html_document
---

The R package dplyr was written to simplify your life.  If you adopt the 
conventions of the package your code will likely be quite a bit neater, 
more efficient, faster and easier to read.  The package is not sliced bread,
but it generally prevents us from adopting numerous syntaxes and styles within a 
single script.

A nice introduction (in the form of a vignette) can be found at [this link.](https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html)

The packages author, Hadley Wickham, likes to discuss the grammar of data.  The 
analogy helps us think about how the package is organized.  We use verbs to do 
things.  

### Main Verbs
1. filter (and slice)
    + filters out the desired values or eliminates undesirable data, i.e. 
    subsetting
    + slice differs in that it allows one to select rows by position (e.g. 34)
    + works on rows
    + base R functions - subset, [
2. arrange
    + orders the data according to the values in one or more columns
    + base R functions - order, sort
3. select (rename)
    + allows the user to select certain columns or delete them using a negative
    sign
    + works on columns
    + base R functions - [, $
4. distinct
    + finds unique values within a column or tbl_df
    + works on a column, columns or and entire tbl_df 
    + base R functions - duplicated, unique
5. mutate (transmute, mutate_each)
    + changes the values within a column or adds a new column
    + transmute is different in that it retains only the modified columns
    + base R functions - replace, [
6. summarise (summarise_each)
    + computes summary statistics over some index or indices, multiple values
    become one
    + logical statements usually work within a column, but the effect is on the
    entire tbl_df
    + base R functions - tapply, loops
7. sample_n (sample_frac)
    + allows subsetting of data randomly by row or fraction of the whole
    + works on the entire tbl_df by selecting certain rows
    + base R functions - sample and [ or subset
8. group_by (rowwise)
    + groups your data by the values in one or more columns
    + rowwise differs in that it groups each row as a unique entity
    + works on columns
    + base R functions - none, but similar things are accomplished using split
    followed by lapply or simply a loop over unique values in a column
9. do
    + used to perform arbitrary computations (e.g. custom function) that can
    return a tbl_df or any other R object, but if the result is not a data frame
    the results will be stored in a list
    + can work on any portion of the data or the entire tbl_df
    + base R functions - typically this type of operation would be done using a
    loop or a member of the apply family
    
In addition to the verbs we also have some helpers.

### Helpers
* n, count, n_distinct, tally
* groups, group_inidices, group_size
* lead, lag
* nth, first, last


#### New Class tbl_df
dplyr introduces a new class of object called a tbl_df.  There is nothing you
absolutely need to know about this class because from the user's perspective it
is pretty well just a data frame with a few extra attributes and different print
method.

#### Databases
dplyr has pretty extensive abilities to connect to databases, but they are 
beyond the scope of this document.  I just want you to be aware that they exist.

## Example

Let us start by creating some data.

```{r}
x <- data.frame(ID = rep(letters[1:5], each = 3),
                Sex = rep(sample(c("M", "F"), 5, replace = T), each = 3),
                Status = 1,
                Year = rep(2011:2015, each = 3))
```

We should discuss what was done above and why it worked.

#### Excercises using dplyr

* Do an operation that returns all the data from before 2013
* Now return all the data in 2014
* Now return the data for Males in 2014
* Return the data for Males in 2014, but report only the number of observations
of each animal
* How many animals of each sex were surveyed in each year
* What are the unique values of the ID column and how many observations per
* Sort the result from the previous step in descending order
* Change all the values of Status to 2 where Sex is Male

#### Excercises part 2
```{r}
xx <- data.frame(ID = rep(letters[1:5], each = 3),
                 Value = c(rnorm(14), NA),
                 Year = rep(c(2010:2012), 5))
```

* What is the mean of Values for each ID
* What is the mean of Value for each ID in 2011
* What is the sum of Value by Year

#### Excercises part 4
In the vignette linked above there are many examples of using dplyr.  Here I 
simply ask the questions that are answered in the vignette.  We are using a 
dataset 'r ?nycflights13' contained in the package nycflights13.  Install and 
then load the package before you try to work with the data.

```r
install.packages("nycflights13")
library(nycflights13)
head(flights)
```

* Subset the data to only those flights on January 1st
* Subset to rows 6 through 9
* Order the data by year, month and day
* Order the data as before, but this time return the only the tail number 
column and change the name to tail_nums
* Now summarise the data by returning only the unique tail numbers
* Find the unique combinations of origin and destination
* Find the difference between the arrival delay and the depature delays
* Do it again, but return a single result for each year
* Randomly select 5 rows from the data

* Return the number of flights by tail number along with the mean distance that
plane travels and the average delay

* Take the above data and subset to only those with more than 20 flights and
flight distance less than 2000
